---
title: "Green Space Landscape Metrics"
author: "Hassan, Danny and Akhil"
format: html
editor: visual
---

# Step 1. Load packages and data

```{r}
library(terra)
library(sf)
library(landscapemetrics)

green <- rast(
  "D:/ARFW0501/newest/Data/green_binary_local_v2.tif")
# Step 1 - change this line:

grid <- st_read(
  "D:/ARFW0501/newest/Data/grid.gpkg",
  quiet = TRUE
)

plot(green)
plot(st_geometry(grid), add = TRUE)
```

# Step 2. Test one grid cell

```{r}
test_cell <- vect(grid[500, ])

green_test <- crop(green, test_cell)

green_test <- mask(green_test, test_cell)

plot(green_test)

```

# Step 3. Calculate landscape metrics one grid cell

```{r}
lsm_p_area(green_test)
lsm_p_gyrate(green_test)
lsm_p_contig(green_test)
lsm_p_enn(green_test)
lsm_p_PROX(green_test)

```

# Step 4. Calculate metrics for all grid cells

```{r}
results <- data.frame(id = grid$id)

results$area <- NA
results$gyrate <- NA
results$contig <- NA
results$enn <- NA
results$prox <- NA

for (i in 1:nrow(grid)) {

  cell <- vect(grid[i, ])

  green_cell <- crop(green, cell)
  green_cell <- mask(green_cell, cell)

  results$area[i] <- mean(lsm_p_area(green_cell)$value[lsm_p_area(green_cell)$class == 1], na.rm = TRUE)

  results$gyrate[i] <- mean(lsm_p_gyrate(green_cell)$value[lsm_p_gyrate(green_cell)$class == 1], na.rm = TRUE)

  results$contig[i] <- mean(lsm_p_contig(green_cell)$value[lsm_p_contig(green_cell)$class == 1], na.rm = TRUE)

  results$enn[i] <- mean(lsm_p_enn(green_cell)$value[lsm_p_enn(green_cell)$class == 1], na.rm = TRUE)

  results$prox[i] <- mean(lsm_p_prox(green_cell)$value[lsm_p_prox(green_cell)$class == 1], na.rm = TRUE)
}

head(results)
```

# Step 5. Join metrics to grid

```{r}
grid_metrics <- merge(
  grid,
  results,
  by = "id"
)
```

# Step 6. contitguity

```{r}
plot(grid_metrics["area"], main = "Mean Patch Area")

```

# Step 7. Division

```{r}
plot(grid_metrics["gyrate"], main = "Mean Radius of Gyration")

```

# Step 8. Aggregation index

```{r}
plot(grid_metrics["contig"], main = "Mean Contiguity Index")

```

# Step 9. Mesh

```{r}
plot(grid_metrics["enn"], main = "Mean Nearest-Neighbor Distance")

```

# Step 10. Mean patch area

```{r}
plot(grid_metrics["prox"], main = "Mean Proximity Index")

```

# Step 11. Load flood raster

```{r}
file.exists("D:/ARFW0501/Delft_Sentinel/Afterminusbefore_floodmask_Delft_28992.tif")
# 1. Check file size - if near 0 bytes it's empty
file.info("D:/ARFW0501/Delft_Sentinel/Afterminusbefore_floodmask_Delft_28992.tif")$size

# 2. Try with GDAL info directly
library(terra)
describe("D:/ARFW0501/Delft_Sentinel/Afterminusbefore_floodmask_Delft_28992.tif")
flood <- rast(
  "D:/ARFW0501/Delft_Sentinel/Afterminusbefore_floodmask_Delft_28992.tif"
)

# Check it loaded correctly
plot(flood, main = "Flood change mask (1 = newly flooded)")
flood  # print metadata - check CRS and resolution match green
```

# Step 12. Align flood raster to green raster CRS

```{r}
# Only reproject if CRS does not match - check first
crs(flood) == crs(green)

# If FALSE, reproject:
#flood <- project(flood, green)
```

# Step 13. Extract flood metrics per grid cell

```{r}

# Step 1: Snap flood to green raster grid exactly
flood_aligned <- resample(flood, green, method = "near")

# Step 2: Classify with NaN handled
flood_binary <- classify(flood_aligned,
  rbind(c(-Inf, 0.05, 0),
        c(0.05,  Inf, 1)),
  others = NA)

# Compare extents
ext(flood_binary)
ext(grid)

# Plot both to see if they overlap
plot(flood_binary, main = "Flood Binary Mask with Grid Overlay (Delft)")
plot(st_geometry(grid), add = TRUE)

# Step 3: Confirm - should show only 0 and 1
freq(flood_binary)
unique(values(flood_binary))

# Set up results data frame
flood_results <- data.frame(id = grid$id)
flood_results$pland_flood <- NA
flood_results$np_flood    <- NA
flood_results$cohesion    <- NA
flood_results$clumpy      <- NA
flood_results$division    <- NA
flood_results$lpi         <- NA

for (i in 1:nrow(grid)) {
  tryCatch({
    cell       <- vect(grid[i, ])
    flood_cell <- crop(flood_binary, cell)
    flood_cell <- mask(flood_cell, cell)

    vals <- na.omit(values(flood_cell))
    if (length(vals) == 0 || length(unique(vals)) < 2) next

    pland    <- lsm_c_pland(flood_cell)
    np       <- lsm_c_np(flood_cell)
    coh      <- lsm_c_cohesion(flood_cell)
    clumpy   <- lsm_c_clumpy(flood_cell)
    division <- lsm_l_division(flood_cell)
    lpi      <- lsm_c_lpi(flood_cell)

    if (nrow(pland[pland$class == 1,]) > 0)
      flood_results$pland_flood[i] <- pland$value[pland$class == 1]
    if (nrow(np[np$class == 1,]) > 0)
      flood_results$np_flood[i]    <- np$value[np$class == 1]
    if (nrow(coh[coh$class == 1,]) > 0)
      flood_results$cohesion[i]    <- coh$value[coh$class == 1]
    if (nrow(clumpy[clumpy$class == 1,]) > 0)
      flood_results$clumpy[i]      <- clumpy$value[clumpy$class == 1]
    if (nrow(division) > 0)
      flood_results$division[i]    <- division$value
    if (nrow(lpi[lpi$class == 1,]) > 0)
      flood_results$lpi[i]         <- lpi$value[lpi$class == 1]

  }, error = function(e) {
    message("Skipping cell ", i, ": ", e$message)
  })
}

print(flood_results)

# Check how many cells actually have flooding
sum(!is.na(flood_results$pland_flood))

# Join green metrics + flood metrics (grid_metrics must exist from Step 5)
grid_full <- merge(grid_metrics, flood_results, by = "id")

# Count cells with both green and flood values
sum(!is.na(grid_full$contig) & !is.na(grid_full$pland_flood))
sum(!is.na(grid_full$area)   & !is.na(grid_full$pland_flood))

# Plots
par(mfrow = c(2, 3))
plot(grid_full["pland_flood"], main = "% Cell Flooded (pland)")
plot(grid_full["np_flood"],    main = "Number of Flood Patches")
plot(grid_full["cohesion"],    main = "Flood Patch Cohesion")
plot(grid_full["clumpy"],      main = "Flood Clumpiness")
plot(grid_full["division"],    main = "Landscape Division")
plot(grid_full["lpi"],         main = "Largest Flood Patch Index")
par(mfrow = c(1, 1))

# Save outputs
st_write(grid_full, 
         "D:/ARFW0501/Delft_Sentinel/grid_full_metrics.gpkg",
         delete_if_exists = TRUE)

writeRaster(flood_binary,
            "D:/ARFW0501/Delft_Sentinel/flood_binary_delft.tif",
            overwrite = TRUE)
```

# Step 14. Join flood metrics to green metrics

```{}
```

# Step 15. Correlate green configuration vs flood extent

```{r}
cor.test(grid_full$contig,  grid_full$pland_flood, use = "complete.obs")
cor.test(grid_full$enn,     grid_full$pland_flood, use = "complete.obs")
cor.test(grid_full$prox,    grid_full$pland_flood, use = "complete.obs")
cor.test(grid_full$area,    grid_full$pland_flood, use = "complete.obs")
```

# Step 16. Plot: green contiguity vs flood extent

```{r}
plot(grid_full$contig, grid_full$pland_flood,
     xlab = "Mean Green Contiguity",
     ylab = "% Cell Flooded",
     main = "Green contiguity vs flood extent per cell (Delft)")
```
