---
title: "Green Space Landscape Metrics"
author: "Hassan, Danny and Akhil"
format: html
editor: visual
---

# Step 1. Load packages and data

```{r}
library(terra)
library(sf)
library(landscapemetrics)

green <- rast(
  "D:/Documents/TU Delft/GEOMATICS/spatial/week 5/Data/qgis/xian/UGS-1m/reprojected_green_cropped_xian_binary.tif"
)

grid <- st_read(
  "D:/Documents/TU Delft/GEOMATICS/spatial/week 5/Data/qgis/grid_10000x10000_xian.gpkg",
  quiet = TRUE
)

plot(green)
plot(st_geometry(grid), add = TRUE)
```

# Step 2. Test one grid cell

```{r}
test_cell <- vect(grid[500, ])

green_test <- crop(green, test_cell)

green_test <- mask(green_test, test_cell)

plot(green_test)

```

# Step 3. Calculate landscape metrics one grid cell

- **Largest Patch Index (LPI):** Measures how much of the landscape is occupied by the largest green patch.

<!-- -->

- **Division Index:** Measures how fragmented the green space is, with higher values indicating greater fragmentation.

<!-- -->

- **Aggregation Index (AI):** Measures how clustered and connected green patches are across the landscape.

<!-- -->

- **Effective Mesh Size (MESH):** Measures the degree of landscape connectivity by estimating the size of connected green areas.

<!-- -->

- **Mean Patch Area (AREA_MN):** Measures the average size of green patches within the grid cell.

<!-- -->

- **Patch Area (AREA):** Measures the size of each individual green patch.

<!-- -->

- **Radius of Gyration (GYRATE):** Measures how far a patch extends from its centre, indicating its spatial reach.

<!-- -->

- **Contiguity Index (CONTIG):** Measures how internally connected and continuous a patch is.

<!-- -->

- **Euclidean Nearest Neighbour Distance (ENN):** Measures the distance to the nearest neighbouring green patch, indicating patch isolation.

```{r}
lsm_c_lpi(green_test)
lsm_c_division(green_test)
lsm_c_ai(green_test)
lsm_c_mesh(green_test)
lsm_c_area_mn(green_test)

lsm_p_area(green_test)
lsm_p_gyrate(green_test)
lsm_p_contig(green_test)
lsm_p_enn(green_test)

```

# Step 4.1 Calculate metrics for all grid cells

```{r}

results_class <- data.frame(id = grid$id)

results_class$lpi <- NA
results_class$division <- NA
results_class$ai <- NA
results_class$mesh <- NA
results_class$area_mn <- NA

for (i in 1:nrow(grid)) {

  cell <- vect(grid[i, ])

  green_cell <- crop(green, cell)
  green_cell <- mask(green_cell, cell)

  results_class$lpi[i] <- lsm_c_lpi(green_cell)$value[1]
  results_class$division[i] <- lsm_c_division(green_cell)$value[1]
  results_class$ai[i] <- lsm_c_ai(green_cell)$value[1]
  results_class$mesh[i] <- lsm_c_mesh(green_cell)$value[1]
  results_class$area_mn[i] <- lsm_c_area_mn(green_cell)$value[1]
}

head(results_class)
```

# Step 4.2 Calculate metrics for all grid cells

```{r}
results_patch <- data.frame(id = grid$id)

results_patch$area <- NA

for (i in 1:nrow(grid)) {

  cell <- vect(grid[i, ])
  green_cell <- crop(green, cell)
  green_cell <- mask(green_cell, cell)

  area <- lsm_p_area(green_cell)

  results_patch$area[i] <- mean(area$value[area$class == 1], na.rm = TRUE)
}

head(results_patch)
```

# Step 4.3 radius of gyration

```{r}
results_patch$gyrate <- NA

for (i in 1:nrow(grid)) {

  cell <- vect(grid[i, ])
  green_cell <- crop(green, cell)
  green_cell <- mask(green_cell, cell)

  gyrate <- lsm_p_gyrate(green_cell)

  results_patch$gyrate[i] <- mean(gyrate$value[gyrate$class == 1], na.rm = TRUE)
}

head(results_patch)
```

# Step 4.4 contiguity index

```{r}
results_patch$contig <- NA

for (i in 1:nrow(grid)) {

  cell <- vect(grid[i, ])
  green_cell <- crop(green, cell)
  green_cell <- mask(green_cell, cell)

  contig <- lsm_p_contig(green_cell)

  results_patch$contig[i] <- mean(contig$value[contig$class == 1], na.rm = TRUE)
}

head(results_patch)
```

# Step 4.5 nearest neighbor distance

```{r}
results_patch$enn <- NA

for (i in 1:nrow(grid)) {

  cell <- vect(grid[i, ])
  green_cell <- crop(green, cell)
  green_cell <- mask(green_cell, cell)

  enn <- lsm_p_enn(green_cell)

  results_patch$enn[i] <- mean(enn$value[enn$class == 1], na.rm = TRUE)
}

head(results_patch)
```

# Step 5.1 Combine class and patch metrics

```{r}
results <- merge(
  results_class,
  results_patch,
  by = "id"
)

head(results)
```

# Step 5.2 Join metrics to grid

```{r}
grid_metrics <- merge(
  grid,
  results,
  by = "id"
)
```

# Step 5.3 Load/calculate FWEI change

```{r}
fwei_before <- rast(
  "D:/Documents/TU Delft/GEOMATICS/spatial/week 5/Data/qgis/wetransfer_fwei_7_22_2023_xian-tif_2026-06-10_1523/CLIPPED_FWEI_7_22_2023_Xian.tif"
)

fwei_after <- rast(
  "D:/Documents/TU Delft/GEOMATICS/spatial/week 5/Data/qgis/wetransfer_fwei_7_22_2023_xian-tif_2026-06-10_1523/CLIPPED_FWEI_8_16_2023_Xian.tif"
)

fwei_change <- fwei_after - fwei_before

plot(fwei_change, main = "FWEI Change Xian")
```

# 

# Step 5.4 extract FWEI to grid

```{r}
fwei_mean <- terra::extract(
  fwei_change,
  vect(grid),
  fun = mean,
  na.rm = TRUE
)

names(fwei_mean)[2] <- "fwei_change_mean"

head(fwei_mean)
```

# 

# Step 5.5 Join FWEI to metrics

```{r}
grid_metrics$fwei_change_mean <- fwei_mean$fwei_change_mean

head(grid_metrics)
```

# 

# Step 6. plotting

```{r}
plot(grid_metrics["lpi"], main = "Largest Patch Index")
plot(grid_metrics["division"], main = "Division Index")
plot(grid_metrics["ai"], main = "Aggregation Index")
plot(grid_metrics["mesh"], main = "Effective Mesh Size")
plot(grid_metrics["area_mn"], main = "Mean Patch Area")
plot(grid_metrics["area"], main = "Patch Area")

```

# 

# Step 7 corelation

| Metric   | Correlation |
|----------|-------------|
| LPI      | -0.106      |
| Division | +0.099      |
| AI       | -0.081      |
| Mesh     | -0.099      |
| Area_MN  | -0.089      |
| Area     | +0.111      |

```{r}
cor(grid_metrics$lpi, grid_metrics$fwei_change_mean, use = "complete.obs")
cor(grid_metrics$division, grid_metrics$fwei_change_mean, use = "complete.obs")
cor(grid_metrics$ai, grid_metrics$fwei_change_mean, use = "complete.obs")
cor(grid_metrics$mesh, grid_metrics$fwei_change_mean, use = "complete.obs")
cor(grid_metrics$area_mn, grid_metrics$fwei_change_mean, use = "complete.obs")
cor(grid_metrics$area, grid_metrics$fwei_change_mean, use = "complete.obs")
```
